A free/open-source Kazakh-Tatar machine translation system

### 2013-04-22 paper/paper.tex
wer numbers

%maybe this should be moved to 'Future plans'
Within the Apertium project, work on several Turkic-language MT systems has been started, with the ultimate goal of creating independent but compatible
finite-state transducers for each language.
%% The idea is that we started Kazakh-Tatar pair to be able to take into account
%% characteristics of two more distant Turkic languages, avoiding narrowing
%% ourselves to mutually intellegible Tatar and Bashkir, but generate
%% transducers which are still compatible and very coherent. Bashkir part of the
%% pair will be updated to the latest developments of Tatar transducer, which
%% took place while working on Kazakh-Tatar.

@@

We had two small Kazakh corporas alongside with their postedited translations into Tatar to measure the WER.
The first one (2457 words total) was a concatenation of an article from Radio Free Europe / Radio Liberty's
Kazakh-language service, an article from Wikipedia, and a simple story used for pedagogical purposes in a workshop
on MT for the languages of Russia. In addition to postediting the translation, this corpus was run through the
morphological transducer and its output was manually disambiguated. All the stems in these texts had been added to
the system, and all the rules (Constraint Grammar,lexical selection and transfer) were based on this corpus.
Hence we can call it the development corpus. It presents an upper bound on the current performance
of the system. Table FIXME http://pastebin.com/TaaWxJzN presents the WER for it.

The second corpus -- a collection of only Radio Free Europe / Radio Liberty's
Kazakh-language service articles -- was used solely for evaluation. It had a similar size -- 2862 words.
The WER results for it can be seen in the table FIXME http://pastebin.com/7tzGCgMX.

@@

\begin{table}[htbp]
  \begin{center}
  \begin{tabular}{ccrrr}
  \toprule
   Corpus                 & Direction         & Tokens  & Unknown & WER  \\
  \midrule
   \multirow{2}{*}{story} & tt$\rightarrow$ba & 311     & 9  & 8.97\% \\
                          & ba$\rightarrow$tt & 312     & 1  & 7.72\%  \\
  \bottomrule
  \end{tabular}
    \caption{Word error rate and over the small test corpus.}
 %\vspace{-2.5em}
    \label{table:wer}
  \end{center}
\end{table}

>>>>

%maybe this should be moved to 'Future plans'
%Within the Apertium project, work on several Turkic-language MT systems has been started, with the ultimate goal of creating independent but compatible
%finite-state transducers for each language.
%% The idea is that we started Kazakh-Tatar pair to be able to take into account
%% characteristics of two more distant Turkic languages, avoiding narrowing
%% ourselves to mutually intellegible Tatar and Bashkir, but generate
%% transducers which are still compatible and very coherent. Bashkir part of the
%% pair will be updated to the latest developments of Tatar transducer, which
%% took place while working on Kazakh-Tatar.

@@

We had two small Kazakh corporas alongside with their postedited translations into Tatar to measure the WER.
The first one (2457 words total) was a concatenation of an article from Radio Free Europe / Radio Liberty's
Kazakh-language service, an article from Wikipedia, and a simple story used for pedagogical purposes in a workshop
on MT for the languages of Russia. In addition to postediting the translation, this corpus was run through the
morphological transducer and its output was manually disambiguated. All the stems in these texts had been added to
the system, and all the rules (Constraint Grammar,lexical selection and transfer) were based on this corpus.
Hence we can call it the development corpus. It presents an upper bound on the current performance
of the system. Table~\ref{table:wer-development} presents the WER for it. % http://pastebin.com/TaaWxJzN

The second corpus -- a collection of only Radio Free Europe / Radio Liberty's
Kazakh-language service articles -- was used solely for evaluation. It had a similar size -- 2862 words.
The WER results for it can be seen in the table~\ref{table:wer-development}. % http://pastebin.com/7tzGCgMX.

@@

\begin{table}[htbp]
  \begin{center}
  \begin{tabular}{ccrrr}
  \toprule
   Corpus                 & Direction           & Tokens  & Unknown & WER  \\
  \midrule
  Development corpus      & kaz$\rightarrow$tat & 2457    & 2       & 15.19\% \\
  \bottomrule
  \end{tabular}
    \caption{Word error rate over the development corpus.}
 %\vspace{-2.5em}
    \label{table:wer-development}
  \end{center}
\end{table}

\begin{table}[htbp]
  \begin{center}
  \begin{tabular}{ccrrr}
  \toprule
   Corpus                 & Direction           & Tokens  & Unknown & WER  \\
  \midrule
  Evaluation corpus       & kaz$\rightarrow$tat & 2862    & 43      & 36.57\% \\
  \bottomrule
  \end{tabular}
    \caption{Word error rate over the evaluation corpus.}
 %\vspace{-2.5em}
    \label{table:wer-evaluation}
  \end{center}
\end{table}
